EuPPollNet: A European database of plant-pollinator networks

### 2024-06-07 Manuscript/EuPPollNet_Ms.qmd
edits

Here, we present the European Plant-Pollinator Networks database (EuPPollNet), which contains harmonized information on interaction data of plants and pollinators at the European level. The pollinator taxonomic groups are composed by the main orders of entomofauna that visit and pollinate flowering plants in Europe. These include insect species from the orders Hymenoptera, Diptera, Lepidoptera, and Coleoptera [@willemstein1987; @potts2015], which comprise almost the totality of recorded interactions in EuPPollNet (`r format(round(main_orders_percent, 2), nsmall = 2)`%). To understand the scope of the database, we examine the taxonomic and sampling coverage of the different plant and pollinator species at European level with the help of the most up-to-date species checklists and rarefaction analyses. In addition, we investigate how two key structural metrics of plant-pollinator networks, connectance and nestedness, change across the latitudes range of studies and bioclimatic regions. Given that this database contains one of the largest sets of plant-pollinator networks collated to date, and considering that the prevalence of nestedness has been evaluated only with a relatively small number of networks [@bascompte2003], we also examine whether plant-pollinator networks are more nested than expected by chance. EuPPollNet aims to cover a wide range of taxonomic groups and habitats, while also providing other variables of interest that define the ecological context and sampling methods of the study. In addition, EuPPollNet offers a transparent and accessible workflow of its data management and species harmonization that allows the database to be reused and to expand over time. This database provides a large number of community-level networks with curated and harmonized data, distinguishing it from other currently available resources that contain plant-pollinator interactions. We expect that EuPPollNet can be used to evaluate macro-ecological processes in plant-pollinator networks, guide conservation planning, and set a baseline for global change research.

@@

The EuPPollNet database is available in both .csv and .rds formats. The .csv file contains data in long format, while the .rds file stores the data in a list structure, with networks organized within each study. The file contains a total of `r nrow(my_tibble) + 1` columns that depict where and when the plant-pollinator interaction was observed (**Table 1**). To provide a standard unit of plant-pollinator interaction across studies, each row depicts a single interaction between a plant and a pollinator species. To build a plant-pollinator network matrix within a single flowering season at a site level, users only need to group interactions by plant and pollinator species, site, study and year. Information about habitat type and bioclimatic region are also provided in this file. In addition, the interaction dataset includes a column that describes the presence or absence of floral counts for each study. The flower count data is provided in a separate file. Two thirds of studies contain information on floral abundance, but the methods and units differ greatly across studies. Finally, metadata at the study level is provided, including information about the authors, digital object identifier if available, sampling time and taxonomic coverage of the main pollinator groups for each study.

@@

For pollinators: (i) we first created a checklist of species names for the most representative pollinator groups at the European level by combining the most up to date published checklists of bees and syrphids [@reverte2023] with the one of butterflies [@wiemers2018]; (ii) we compared pollinator species names against the checklist and recovered some unmatched cases with restrictive fuzzy matching by using **stringdist** package [@van2014]; (iii) we fixed unmatched records when necessary and retrieved the taxonomic information for all species from GBIF; (iv) we fixed the non-found cases in the GBIF checklist and made sure that all species names from bees, syrphids and butterflies were named according to their respective species checklists. Coleoptera species names were only checked against the GBIF checklist.

@@

To assess the completeness of plant and pollinator species in the EuPPollNet database at European level, we used the aforementioned checklists for plants and pollinators. Specifically for plants, we refined the checklist to include only European flowering plants and excluded taxonomic groups not associated with biotic pollination. We did this by first excluding the families considered to have exclusively a wind pollination mode [see @culley2002], and then by filtering out the genera with wind or non-biotic pollination from families that exhibit both biotic and non-biotic pollination modes. Additionally, we manually included some exotic species and added unresolved species names that were not present in the accepted names of the checklist at the current version of usage. For pollinators, we compared only the taxonomic coverage of bees, syrphids and butterflies by using their species checklists at European level. The potential number of pollinator species at European level was estimated by adding the total number of species of bees, syrphids, and flies from the checklists, along with the extrapolated number of species from other insect pollinators. This extrapolated number was estimated by assuming that the sampling coverage of these other insect pollinators is equal to the average coverage across bees, syrphids, and butterflies (mean coverage = `r round(mean_coverage, 2)`; sd = `r round(sd_coverage,2)`).

@@

The different sites were described with a habitat type by the authors of each study. We standardized  the habitat type using the author's description and land cover information from Corine Land Cover (CLC, version 2018) extracted using the Terra package [@hijmans2022]. Based on the habitat classification from the authors and the CLC classification, we created habitat categories that intend to summarize the diversity of habitats in the EuPPollNet database (see habitat type definition in supplementary material). These categories allow a quick comparison and understanding of the habitat types from the database. Moreover, Europe is characterized by a great variety of environmental conditions that harbor different biota. Thus, to allow authors to explore the set of studies that share similar environmental conditions and species, we assigned to each site a biogeographical region. The biogeographical regions were downloaded from the European Environment Agency (version 2016) and were matched to the different sites using a spatial joint from the **sf** package [@pebesma2018].

@@

To provide a general overview of the structure of plant-pollinator networks in EuPPollNet, we quantified connectance and nestedness and examined how these network metrics change across the latitudinal range of studies and bioclimatic regions. We selected these two network metrics as they are commonly evaluated in plant-pollinator network studies and capture structural properties with a straightforward interpretation. However, these network metrics are not independent of sampling effort (e.g., number of species) and to allow comparisons across them, we implemented “standardised” versions of connectance and nestedness. As connectance is negatively associated with network size [@jordano1987], we evaluated how network connectance was associated with the number of species  (i.e., log of geometric mean of plants and pollinators) and extracted the residuals from this association (i.e., residual connectance) as a measurement of corrected connectance. This was done with the help of a Beta regression implemented with the package **betareg** [@cribari2010]. Second, we implemented a metric of nedstedness (i.e., NODFc) that allows the comparison across networks as it corrects by connectance and the number of species [@song2017]. This metric was calculated with the help of the **maxnodf** package [@hoeppke2021], which employs a computational eﬀicient approach to calculate it. Both residual connectance and NODFc were used as dependent variables to evaluate their association with latitude. In addition, we explored the association between connectance and nestedness with the number of species per network by conducting Kendall rank correlation coefficient in order to be able to compare the strength of this association across both network metrics.

Finally, to compare if networks are more or less nested than expected by chance, we employed the traditional z-score approach with the widely used nestedness metric (NODF) from @almeida2008, as this approach allow us to compare our results with previous published nestedness analysis in plant-pollinator networks and only compares each unique network against their randomized versions. We calculated 100 null models for each network with the help of the _vaznull_ function from the package **bipartite** [@dormann2008] that implements the null model from @vazquez2005. These null networks have the same connectance and number of plant and pollinator species as the empirical ones, but different marginal totals. Both connectance and nestedness (NODF) were estimated for each network by using the function _networklevel_ from **bipartite**.

@@

The proportion of species from the major pollinator orders within the database differed across habitats and bioclimatic regions (**Figure 5**). As expected, Hymenoptera was the main taxonomic order on the majority of habitats, exceeded only by Diptera on the habitat categories of riparian vegetation, moors and heathland, and alpine grasslands. Overall, the proportion of Lepidoptera and Coleoptera species was low across all habitats but Coleoptera showed a notably increase in sclerophyllous vegetation and beaches, dunes and sands habitat categories. Similar patterns were observed when exploring the pollinator proportions by bioclimatic region. In this particular case, Hymenoptera was the predominant order in all bioclimatic regions with Diptera taking more importance in Alpine and Atlantic regions. Lepidoptera shows low proportions across all bioclimatic regions and Coleoptera is only relevant in Mediterranean regions at European level. Notably, the number of studies (**Figure 5**) and sampling sites (**Figure S2**) also differs across habitats and bioclimatic regions. The habitats sampled by a higher number of studies in the database are `r studies_by_habitat$EuPPollNet_habitat[[1]]` (`r studies_by_habitat$Habitat_count[[1]]`), `r studies_by_habitat$EuPPollNet_habitat[[2]]` (`r studies_by_habitat$Habitat_count[[2]]`) and `r studies_by_habitat$EuPPollNet_habitat[[3]]` (`r studies_by_habitat$Habitat_count[[3]]`). However, the habitats that contain a higher number of sampling sites are `r networks_by_habitat$EuPPollNet_habitat[[1]]` (`r networks_by_habitat$Habitat_count[[1]]`), `r networks_by_habitat$EuPPollNet_habitat[[2]]` (`r networks_by_habitat$Habitat_count[[2]]`) and `r networks_by_habitat$EuPPollNet_habitat[[3]]` (`r networks_by_habitat$Habitat_count[[3]]`). The bioclimatic regions with a higher number of studies are `r studies_by_bioregion$Bioregion[[1]]` (`r studies_by_bioregion$Bioregion_count[[1]]`), `r studies_by_bioregion$Bioregion[[2]]` (`r studies_by_bioregion$Bioregion_count[[2]]`) and `r studies_by_bioregion$Bioregion[[3]]` (`r studies_by_bioregion$Bioregion_count[[3]]`); and the ones that contain a higher number of sampling sites are `r networks_by_bioregion$Bioregion[[1]]` (`r networks_by_bioregion$Bioregion_count[[1]]`), `r networks_by_bioregion$Bioregion[[2]]` (`r networks_by_bioregion$Bioregion_count[[2]]`) and `r networks_by_bioregion$Bioregion[[3]]` (`r networks_by_bioregion$Bioregion_count[[3]]`).

@@

Bees are responsible for the majority of the interactions at the metaweb level, but their relative relevance changed across habitats and bioclimatic regions. For instance, plant-pollinator communities in the Mediterranean were dominated by bees, while communities in Alpine or Boreal regions were fly species rich or fly-dominated. These patterns are consistent with our current understanding of bee diversity, which peaks in dry or temperate areas [@orr2021; @leclercq2023]; and with the fact that colder environments (i.e., altitude and latitude wise) harbor a larger fraction of fly pollinators compared to other taxa [@elberling1999; @lefebvre2018]. In addition, beetles were only common floral visitors in the Mediterranean region. Although the networks from the database are visitation networks and do not capture pollinator efficiency [@ballantyne2015], the high proportion of beetles as floral visitors in the Mediterranean provides further support for their potential role as pollinators in this region [@herrera2019; @leon2022]. The number of butterfly species and interactions were relatively low compared to the other taxa. While Europe contains fewer butterfly species than other regions of the world [@ollerton2017], their relevance as pollinators is likely underestimated within this database. This is because a large fraction of studies (~40%) did not sample butterflies, and conventional sampling methods for monitoring other insect pollinators (e.g., bees or flies) may be inadequate for sampling plant-butterfly interactions [@isaac2011]. On the contrary, honey bees were present in `r round(apis_percent_networks, 1)`% of networks and conducted on average a third of the total interactions per network. Indeed, this proportion is higher than the one found when considering only natural communities at a global scale [~13%\; @hung2018], which highlights their key role as potential pollinators of many flowering plants. However, the rise of beekeeping in Europe has led to high honey bee densities, which can negatively impact wild pollinators and the plants they visit [@steffan2000; @magrach2017; @herrera2020], emphasising the need to evaluate these impacts and implement actions if necessary.

Although Europe contains a much larger number of flowering plants than pollinator species (~4 to 1 ratio), the observed number of pollinator species almost doubled the one of plants in the database. This could be explained by the fact that all networks are phytocentric, resulting in sampling bias towards pollinator species [@jordano2016; @vizentin2018]. While animal-centred sampling is likely to increase the plant:pollinator species ratio [e.g., @encinas2023], the spatial scale and environmental context of the sampled communities will also influence their observed diversity, especially given the ability to move of pollinators and the sessile nature of plants. In addition, we found that the accumulation curve of pollinators per plant species does not saturate, which indicates low redundancy of pollinators and that many are regionally “rare”. Rare pollinators have been proven to be functionally important for plant species at large scales [@winfree2018; @simpson2022], highlighting the need to conduct further sampling events to observe these rare species and to effectively understand and protect plant-pollinator biodiversity.

Consistent with @olesen2002, we find that residual connectance (i.e., the deviation from the expected connectance for a given network size) was lower at higher latitudes, while normalised nestedness, which is negatively correlated with residual connectance, increased towards higher latitudes. Networks at lower latitudes in Europe are exposed to higher temperatures and are bee-dominated, which can result in higher visitation rates [@arroyo1985; @classen2015; @herrera2019], and the overall level of pollinator generalization is known to be higher at lower latitudes [@schleuning2012]. These factors should increase the possible number of connections that can be established between plants and pollinators for a given network size, resulting in more connected networks but less nested ones, as these networks are likely to deviate more from a perfect nested or hierarchical structure compared to networks at higher latitudes. Finally, while one third of networks were more nested than expected by chance, two thirds did not show statistical difference with null expectations. Therefore, while this supports the idea that plant-pollinator networks tend to be nested [@bascompte2003], this result suggests that nestedness could be a less prevalent feature than previously thought for plant-pollinator networks [@payrato2019].

Despite this database covering a wide range of habitats across `r country_number` countries, it contains geographical biases that can impact our understanding of plant-pollinator communities [@hughes2021]. For instance, most plant-pollinator networks belong to central Europe, while Eastern Europe and the Mediterranean region are under-represented. This is consistent with previous studies which also report lack of plant-pollinator data for those regions [@bennett2018; @marshall2024], highlighting that this database shows existing patterns in data availability despite the absence of a systematic search for studies. This is especially relevant as Eastern Europe currently has vast landscapes of high quality semi-natural grasslands but is experiencing rapid land use change [@sutcliffe2015], and the Mediterranean region is likely to be severely impacted by climate change [@jaworski2022; @duchenne2020; @pareja2023]. These areas are well known for their rich pollinator diversity [@milivcic2018; @reverte2023], and their under-representation is likely contributing to the low taxonomic coverage of this database at the European level. Although some of the most well studied countries in Europe (e.g., Belgium, The Netherlands) have already experienced land use change and biodiversity loss at the end of the 20th century [@carvalheiro2013], plant-pollinator communities in Europe and across the globe still face current and future threats from climate change [@bartomeus2011; @duchenne2020], land use change [@reidsma2006; @batary2015], or the introduction of alien species [@vila2009; @vanbergen2018]. Therefore, continuous monitoring programs are needed in order to evaluate spatio-temporal changes of species and their interactions across different European habitats and regions. This will allow local and large scale analyses of the status and trends of plant-pollinator communities, effectively informing management and conservation actions.

>>>>

Here, we present the European Plant-Pollinator Networks database (EuPPollNet), which contains harmonized information on interaction data of plants and pollinators at the European level. The pollinator taxonomic groups are composed by the main orders of entomofauna that visit and pollinate flowering plants in Europe. These include insect species from the orders Hymenoptera, Diptera, Lepidoptera, and Coleoptera [@willemstein1987; @potts2015], which comprise almost the totality of recorded interactions in EuPPollNet (`r format(round(main_orders_percent, 2), nsmall = 2)`%). To understand the scope of the database, we examine the taxonomic and sampling coverage of the different plant and pollinator species at European level with the help of the most up-to-date species checklists and rarefaction analyses. The overall number of potential pollinators and flowering plants in Europe was estimated with the help of the different checklists and sampling coverage. Finally, for bees and plants, we evaluated if there is a phylogenetic signal in the presence-absence of interaction data.

JUSTIFY WITH A SENTENCE HERE:
Check that the poll and flower number info is in results! and the phylo signal too.

In addition, we investigate how two key structural metrics of plant-pollinator networks, connectance and nestedness, change across the latitude range of studies and bioclimatic regions. Given that this database contains one of the largest sets of plant-pollinator networks collated to date, and considering that the prevalence of nestedness has been evaluated only with a relatively small number of networks [@bascompte2003], we also examine whether plant-pollinator networks are more nested than expected by chance. EuPPollNet aims to cover a wide range of taxonomic groups and habitats, while also providing other variables of interest that define the ecological context and sampling methods of the study. In addition, EuPPollNet offers a transparent and accessible workflow of its data management and species harmonization that allows the database to be reused and to expand over time. This database provides a large number of community-level networks with curated and harmonized data, distinguishing it from other currently available resources that contain plant-pollinator interactions. We expect that EuPPollNet can be used to evaluate macro-ecological processes in plant-pollinator networks, guide conservation planning, and set a baseline for global change research.

@@

The EuPPollNet database is available in both .csv and .rds formats. The .csv file contains data in long format, while the .rds file stores the data in a list structure, with networks organized within each study. The file contains a total of `r nrow(my_tibble) + 1` columns that depict where and when the plant-pollinator interaction was observed (**Table 1**). Each row depicts a single interaction between a plant and a pollinator species. To build a plant-pollinator network matrix within a single flowering season at a site level, users only need to group interactions by plant and pollinator species, site, study and year. Information about habitat type and bioclimatic region are also provided in this file. In addition, the interaction dataset includes a column that describes the presence or absence of floral counts for each study. The flower count data is provided in a separate file (.csv or .rds) that can be merged with the interaction data through the "Flower_data_merger" column. Two thirds of studies contain information on floral abundance, but the methods and units differ greatly across studies. Finally, metadata at the study level is provided, including information about the authors, digital object identifier if available, sampling time and taxonomic coverage of the main pollinator groups for each study.

@@

For pollinators: (i) we first created a checklist of species names for the most representative pollinator groups at the European level by combining the most up to date published checklists of bees and syrphids [@reverte2023], and butterflies [@wiemers2018]; (ii) we compared pollinator species names against the checklist and recovered some unmatched cases with restrictive fuzzy matching by using **stringdist** package [@van2014]; (iii) we fixed unmatched records when necessary and retrieved the taxonomic information for all species from GBIF; (iv) we fixed the non-found cases in the GBIF checklist and made sure that all species names from bees, syrphids and butterflies were named according to their respective species checklists. Coleoptera species names were only checked against the GBIF checklist.

@@

To assess the completeness of plant and pollinator species in the EuPPollNet database at the European level, we used the aforementioned checklists for plants and pollinators. Specifically for plants, we refined the checklist to include only European flowering plants and excluded taxonomic groups not associated with biotic pollination. We did this by first excluding the families considered to have exclusively a wind pollination mode [see @culley2002], and then by filtering out the genera with wind or non-biotic pollination from families that exhibit both biotic and non-biotic pollination modes. Additionally, we manually included some exotic species and added unresolved species names that were not present in the accepted names of the checklist at the current version of usage. For pollinators, we compared only the taxonomic coverage of bees, syrphids and butterflies by using their species checklists at the European level. The potential number of pollinator species at the European level was estimated by adding the total number of species of bees, syrphids, and flies from the checklists, along with the extrapolated number of species from other insect pollinators. This extrapolated number was estimated by assuming that the sampling coverage of these other insect pollinators is equal to the average coverage across bees, syrphids, and butterflies (mean coverage = `r round(mean_coverage, 2)`; sd = `r round(sd_coverage,2)`).

@@

We describe the habitat type for each site using information from Corine Land Cover (CLC, version 2018) extracted using the Terra package [@hijmans2022], visual inspection of Google Earth imagery and the habitat classification from the authors. These different habitat categories (see definitions in supplementary text) allow a quick comparison and understanding of the habitat types from the database. Moreover, Europe is characterized by a great variety of environmental conditions that harbor different biota. Thus, to allow authors to explore the set of studies that share similar environmental conditions and species, we assigned a biogeographical region to each site. The biogeographical regions were downloaded from the European Environment Agency (version 2016) and were matched to the different sites using a spatial joint from the **sf** package [@pebesma2018].

@@

To provide a general overview of the structure of plant-pollinator networks in EuPPollNet, we quantified connectance and nestedness and examined how these network metrics change across the latitudinal range of studies and bioclimatic regions. We selected these two network metrics as they are commonly evaluated in plant-pollinator network studies and capture structural properties with a straightforward interpretation. We implemented “standardised” versions of connectance and nestedness to account for the effect of sampling effort on network metrics. As connectance is negatively associated with network size [@jordano1987], we evaluated how network connectance was associated with the number of species (i.e., log of geometric mean of plants and pollinators) and extracted the residuals from this association (i.e., residual connectance) as a measurement of corrected connectance. The relationship between residual connectance and species richness was investigated using a beta-regression implemented with the package **betareg** [@cribari2010]. Second, we used NODFc to compare nestedness across networks, as it corrects by connectance and the number of species [@song2017]. This metric was calculated using the **maxnodf** package [@hoeppke2021]. Both residual connectance and NODFc were used as dependent variables to evaluate their association with latitude. In addition, we explored the association between connectance and nestedness with the number of species per network using a Kendall rank correlation coefficient to compare the strength of associations between network structures and species richness across both network metrics.

Finally, to compare if networks are more or less nested than expected by chance, we employed the traditional z-score approach with the widely used nestedness metric (NODF) from @almeida2008. The z-score approach allows us to compare our results with previous published nestedness analyses in plant-pollinator networks and only compares each unique network against their randomized versions. We calculated 100 null models for each network with the help of the _vaznull_ function from the package **bipartite** [@dormann2008]. These null networks have the same connectance and number of plant and pollinator species as the empirical ones, but different marginal totals. Both connectance and nestedness (NODF) were estimated for each network by using the function _networklevel_ from **bipartite**.

@@

The proportion of species from the major pollinator orders within the database differed across habitats and bioclimatic regions (**Figure 5**). As expected, Hymenoptera was the main taxonomic order on the majority of habitats, exceeded only by Diptera on the habitat categories of riparian vegetation, moors and heathland, and alpine grasslands. Overall, the proportion of Lepidoptera and Coleoptera species was low across all habitats but Coleoptera were notably more abundant in sclerophyllous vegetation and beaches, dunes and sands habitat categories. Similar patterns were observed when exploring the pollinator proportions by bioclimatic region. In this particular case, Hymenoptera was the predominant order in all bioclimatic regions with Diptera taking more importance in Alpine and Atlantic regions. Lepidoptera shows low proportions across all bioclimatic regions and Coleoptera is only relevant in Mediterranean regions at European level. Notably, the number of studies (**Figure 5**) and sampling sites (**Figure S2**) also differs across habitats and bioclimatic regions. The habitats sampled by a higher number of studies in the database are `r studies_by_habitat$EuPPollNet_habitat[[1]]` (`r studies_by_habitat$Habitat_count[[1]]`), `r studies_by_habitat$EuPPollNet_habitat[[2]]` (`r studies_by_habitat$Habitat_count[[2]]`) and `r studies_by_habitat$EuPPollNet_habitat[[3]]` (`r studies_by_habitat$Habitat_count[[3]]`). However, the habitats that contain a higher number of sampling sites are `r networks_by_habitat$EuPPollNet_habitat[[1]]` (`r networks_by_habitat$Habitat_count[[1]]`), `r networks_by_habitat$EuPPollNet_habitat[[2]]` (`r networks_by_habitat$Habitat_count[[2]]`) and `r networks_by_habitat$EuPPollNet_habitat[[3]]` (`r networks_by_habitat$Habitat_count[[3]]`). The bioclimatic regions with a higher number of studies are `r studies_by_bioregion$Bioregion[[1]]` (`r studies_by_bioregion$Bioregion_count[[1]]`), `r studies_by_bioregion$Bioregion[[2]]` (`r studies_by_bioregion$Bioregion_count[[2]]`) and `r studies_by_bioregion$Bioregion[[3]]` (`r studies_by_bioregion$Bioregion_count[[3]]`); and the ones that contain a higher number of sampling sites are `r networks_by_bioregion$Bioregion[[1]]` (`r networks_by_bioregion$Bioregion_count[[1]]`), `r networks_by_bioregion$Bioregion[[2]]` (`r networks_by_bioregion$Bioregion_count[[2]]`) and `r networks_by_bioregion$Bioregion[[3]]` (`r networks_by_bioregion$Bioregion_count[[3]]`).

@@

Bees are responsible for the majority of the interactions at the metaweb level, but their relevance for network topology changed across habitats and bioclimatic regions. For instance, plant-pollinator communities in the Mediterranean were dominated by bees, while communities in Alpine or Boreal regions were fly species rich or fly-dominated. These patterns are consistent with our current understanding of bee diversity, which peaks in dry or temperate areas [@orr2021; @leclercq2023]; and with the fact that colder environments (i.e., altitude and latitude wise) harbor a larger fraction of fly pollinators compared to other taxa [@elberling1999; @lefebvre2018]. In addition, beetles were only common floral visitors in the Mediterranean region. Although the networks from the database are visitation networks and do not capture pollinator efficiency [@ballantyne2015], the high proportion of beetles as floral visitors in the Mediterranean provides further support for their potential role as pollinators in this region [@herrera2019; @leon2022]. The number of butterfly species and interactions were relatively low compared to the other taxa. While Europe contains fewer butterfly species than other regions of the world [@ollerton2017], their relevance as pollinators is likely underestimated within this database. This is because a large fraction of studies (~40%) did not sample butterflies, and conventional sampling methods for monitoring other insect pollinators (e.g., bees or flies) may be inadequate for sampling plant-butterfly interactions [@isaac2011]. On the contrary, honey bees were present in `r round(apis_percent_networks, 1)`% of networks and conducted on average a third of the total interactions per network. The proportion of honey bees in networks across Europe is higher than in natural communities across the world [~13%\; @hung2018], potentially reflecting the dominance of pasture habitats in EuPPollNet, the highly generalised nature of honeybees, the important role that honeybees are playing as pollinators in Europe, and/or the rise of beekeeping in Europe [@steffan2000; @magrach2017; @herrera2020].

Although Europe contains a much larger number of flowering plants than pollinator species (~4 to 1 ratio), the observed number of pollinator species in the database was almost double that of the plants. This could be explained by the fact that all networks are phytocentric, resulting in sampling bias towards pollinator species [@jordano2016; @vizentin2018]. While animal-centerred sampling is likely to increase the plant-pollinator species ratio [e.g., @encinas2023], the spatial scale and environmental context of the sampled communities will also influence their observed diversity, especially given the ability to move of pollinators and the sessile nature of plants. In addition, we found that the accumulation curve of pollinators per plant species does not saturate, which indicates low redundancy of pollinators and that many are regionally “rare”. Rare pollinators are functionally important for plant species at large scales [@winfree2018; @simpson2022], highlighting the need to conduct further sampling events to observe these rare species and to effectively understand and protect plant-pollinator biodiversity.

Consistent with @olesen2002, we find that residual connectance (i.e., the deviation from the expected connectance for a given network size) was lower at higher latitudes, while normalised nestedness increased towards higher latitudes. Networks at lower latitudes in Europe are exposed to higher temperatures and are bee-dominated, which can result in higher visitation rates [@arroyo1985; @classen2015; @herrera2019], and the overall level of pollinator generalization is known to be higher at lower latitudes [@schleuning2012]. These factors should increase the possible number of connections that can be established between plants and pollinators for a given network size, resulting in more connected and less nested networks, as these networks are likely to deviate more from a perfect nested or hierarchical structure compared to networks at higher latitudes. Finally, while one third of networks were more nested than expected by chance, two thirds did not show statistical difference with null expectations. Therefore, while this supports the idea that plant-pollinator networks tend to be nested [@bascompte2003], this result suggests that nestedness could be a less prevalent feature than previously thought for plant-pollinator networks [@payrato2019].

Despite this database covering a wide range of habitats across `r country_number` countries, it contains geographical biases that can impact our understanding of plant-pollinator communities [@hughes2021]. For instance, most plant-pollinator networks are sampled from central Europe, while Eastern Europe and the Mediterranean region are underrepresented. This is consistent with previous studies which also report lack of plant-pollinator data for those regions [@bennett2018; @marshall2024], highlighting that this database shows existing patterns in data availability despite the absence of a systematic search for studies. This lack of data is especially relevant for Eastern Europe which has vast landscapes of high quality semi-natural grasslands but is experiencing rapid land use change [@sutcliffe2015], and the Mediterranean region is likely to be severely impacted by climate change [@jaworski2022; @duchenne2020; @pareja2023]. These areas are well known for their rich pollinator diversity [@milivcic2018; @reverte2023], and their under-representation is likely contributing to the low taxonomic coverage of this database at the European level. Although some of the most well studied countries in Europe (e.g., Belgium, The Netherlands) have already experienced land use change and biodiversity loss at the end of the 20th century [@carvalheiro2013], plant-pollinator communities in Europe and across the globe still face current and future threats from climate change [@bartomeus2011; @duchenne2020], land use change [@reidsma2006; @batary2015], or the introduction of alien species [@vila2009; @vanbergen2018]. Therefore, continuous monitoring programs are needed in order to evaluate spatio-temporal changes of species and their interactions across different European habitats and regions. This will allow local and large scale analyses of the status and trends of plant-pollinator communities, effectively informing management and conservation actions.
